\documentclass[11pt,letterpaper]{letter}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{url}

\signature{Ihor Kendiukhov\\
Department of Computer Science\\
University of Tuebingen\\
Sand~14, 72076 Tuebingen, Germany\\
kenduhov.ig@gmail.com}

\address{Ihor Kendiukhov\\
Department of Computer Science\\
University of Tuebingen\\
Sand~14, 72076 Tuebingen, Germany}

\begin{document}

\begin{letter}{Editorial Board\\
PLOS Computational Biology}

\opening{Dear Editors,}

I am resubmitting the manuscript entitled ``Causal intervention validation
of gene regulatory signals in scGPT'' (previous manuscript ID:
PCOMPBIOL-D-26-00541) for consideration as a Research Article in
\textit{PLOS Computational Biology}. I am grateful to the editor and
reviewers for the constructive feedback across both review rounds and
for the opportunity to resubmit.

\textbf{Principal revision: quantitative comparison against previous
methods.} The key comment in the most recent review was that, in addition
to describing the existing literature, the manuscript must
\textit{quantitatively} compare the proposed method against previous
methods. This second revision addresses that comment directly by adding a
new Results subsection, \textit{Comparison with classical GRN inference
methods}, a new table, and a new figure that report AUPR, AUROC, and
permutation $p$-values for the scGPT causal ablation alongside eleven
classical GRN baselines spanning six algorithmic families on a matched
reproduction of the main kidney, lung, and immune substrates. The
baselines are: Pearson, Spearman, and distance correlation (coexpression);
partial correlation on a ridge-regularized precision matrix (Gaussian
graphical model / PC-algorithm family); pairwise mutual information, CLR
(Faith et al.\ 2007), and ARACNE (Margolin et al.\ 2006)
(information-theoretic family); LASSO (Tibshirani 1996) and elastic net
(Zou \& Hastie 2005) regression (the Inferelator lineage); and GRNBoost2
(Moerman et al.\ 2019) and GENIE3 (Huynh-Thu et al.\ 2010) (tree-based
regression family). Every baseline is implemented from scikit-learn,
SciPy, and \texttt{dcor} primitives and runs CPU-only, so the entire
comparison is self-contained and reproducible end-to-end from the raw
h5ad files, without any dedicated cluster or GPU backend.

Each baseline is evaluated on a matched reproduction of the main-run
substrate: same raw Tabula Sapiens h5ad files, same preprocessing recipe
(normalize\_total~=~10\,000, log1p, 5\,000 highly variable genes
restricted to the scGPT vocabulary), 120-cell sample drawn with seed 42
(\texttt{max\_cells=120} in the main config), deterministic 120 positive
+ 120 random negative TRRUST pair sampling from the same pair-building
logic as \texttt{run\_causal\_interventions.py}, and the same top-1200
evidence filter used by the scGPT dataset. On this matched substrate, the
scGPT causal ablation AUPR exceeds the strongest classical baseline by
+0.053 in kidney (vs.\ Spearman correlation, 0.549), +0.075 in lung
(vs.\ ridge-regularized partial correlation, 0.652, permutation
$p{=}0.001$), and +0.002 in immune (vs.\ Spearman correlation, 0.594).
The tissue ordering mirrors the cross-tissue pattern the paper already
reports from the main runs: lung is the tissue where both the
permutation-significant enrichment and the scGPT-versus-classical gap
are largest, kidney sits between, and immune is effectively a tie,
matching the paper's observation that immune causal scores sit close to
attention and coexpression baselines.

Two auxiliary observations from the benchmark are discussed in the new
Results and Discussion paragraphs. First, the strongest classical
competitors are correlation-based methods (partial correlation in lung,
Spearman correlation in kidney and immune) rather than the more
computationally elaborate tree-based or information-theoretic methods,
which sit near chance at the 120-cell budget used throughout the paper.
This is consistent with BEELINE (Pratapa et al.\ 2020), which documents
that tree-based and information-theoretic GRN inference typically
requires many hundreds to thousands of cells to converge on direct
edges. Second, the comparison therefore speaks to signal efficiency at
modest cell counts rather than to large-scale absolute accuracy, which
the existing GRN benchmarking literature already covers in depth: a
pretrained foundation model that has already absorbed tens of millions
of cells can provide directional regulatory evidence on a small held-out
sample where classical methods do not yet have enough observations, while
the classical methods remain the better choice when large cell numbers
are available and the goal is explicit network construction rather than
interrogation of a pretrained model's internal representations. These
framings are stated explicitly in the Discussion.

\textbf{Previous review round.} The editor's two previous comments on
the first revision concerned the qualitative positioning of the method
relative to existing GRN inference approaches and prior foundation-model
evaluations. Those comments were addressed by an expanded literature
survey in the Introduction (from two short paragraphs to four, adding 19
new validated references) and by expanded positioning paragraphs in the
Discussion. The present second revision retains all of that material and
adds the quantitative comparison on top of it, so the manuscript now
presents both the conceptual distinction and the empirical head-to-head
numbers the reviewer asked for.

\textbf{Summary of changes in this revision.} New Results subsection
\textit{Comparison with classical GRN inference methods}; new table
reporting AUPR, AUROC, and permutation $p$-values for scGPT plus eleven
classical baselines across kidney, lung, and immune; new grouped-bar
figure visualizing the same comparison; updated Abstract, Introduction,
Discussion, and Conclusion to mention the comparison and its findings;
three new references (Tibshirani 1996, Zou \& Hastie 2005, Sz\'ekely et
al.\ 2007); and two new reproducibility scripts
(\texttt{scripts/grn\_baseline\_comparison.py} and
\texttt{scripts/plot\_grn\_baseline\_comparison.py}) together with
per-tissue metric TSVs under \texttt{outputs/grn\_baseline\_comparison/}.
A detailed point-by-point response, including the full
matched-substrate table, accompanies this resubmission in
\textit{response\_to\_reviewers.pdf}.

Source code, evaluation artifacts, scripts, output TSVs, and the updated
manuscript are publicly available at
\url{https://github.com/Biodyn-AI/casual-intervention}.

This manuscript has not been published elsewhere and is not under
consideration at any other journal. No conflicts of interest exist. I
declare that large language models were used for writing assistance and
code generation during the preparation of this manuscript.

Thank you for considering this resubmission.

\closing{Sincerely,}

\end{letter}
\end{document}
